---
title: "Problem Set - 4"
author: "Aravind K Mannarswamy"
date: "December 02, 2020"
output:
  pdf_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    number_sections: yes
    toc: yes
    toc_depth: 3

---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
library(DT)
library(knitr)
```


# Value of Statistical Life(VSL)   


## What is the VSL?   

Annual chance of death from heart attack is 0.2% or 2 in every 1000. This translates to 20 in every 10,000. A reduction of 5% in this number would be exactly 1.  

So the VSL of an individual is = 10,000 people paying \$500 = \$5,000,000.   
Alternatively, 500/(0.05*0.002) = \$5,000,000. 
    
## Re-do the survey   
  
### VSL estimates of people who have already had a heart attack    
Suppose we want to estimate the cost of treatment that could potentially save someone from heart attack who is already at high risk of dying from it should use this approach.Another hypothetical example could be to determine the cost of treatment to reverse type-2 diabetes.  


### VSL estimates of entire population   
Vaccination of the entire population for seasonal flu or a pandemic is a classical example of when VSL estimates of the entire population should be used.  

# Quality of Life(Artificial Retina)    

```{r, echo=FALSE, message=FALSE, warning=FALSE}
df2 = data.frame(groups = c('sighted','became_blind_long','recently_blinded'),
                 quality = c(0.4,0.1,0.9),
                 stringsAsFactors = FALSE)

df2 %>% 
  ggplot()+
  aes(x = groups, y = quality, color = groups, fill = groups)+
  geom_point(size = 10, alpha = 0.5)+
  theme_bw()+
  theme(legend.position = "none",
        panel.grid = element_blank())+
  scale_x_discrete(labels = c("Recently Became Blind","Blinded long time ago","Sighted People"))+
  scale_y_continuous(labels = c(0.1,0.4,0.9), breaks = c(0.1,0.4,0.9))+
  ylab("Quality of life")
```

## Why are the numbers different for different approaches   

1. All the three groups have different baselines for their 'subjective' evaluation of the quality life.   
2. People who have lived with blindness for 20 years may have already accustomed to being blind as their baseline and insteadwill focus on health ailments to determine their quality of life. In contrast, people who recently became blind has their baseline set at being sighted. This difference in where they were 'recently' must have affected their decision to rate as such.   
   
## Appropriate approach for a possible side effect   
Using q = 0.4, that is the people who are sighted should be the right number to be used. The reason is that the side effect of the drug is becoming blind, that calls for the assumption that someone who can see could potentially become blind due to the use of the drug.   
   
## Appropriate approach for restoring sight to the blind   
Using either q = 0.1 or using q = 0.9, depending on the patient should be fair. However, a weighted average of these responses should be the right approach. For example, if the survey had `n1` long time blind and `n2` recently became blind people, then the appropriate weighted measure would be $(0.9 n_1 + 0.1n_2)/(n_1 + n_2)$      


